Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
subcritically -> sub-critically
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@895 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: Many small fixes and fill-ins
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@906 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in a
search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain as an \textsc{fir} filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point operations per sample.  However, it is typically much more computationally efficient to use the convolution theorem and the \textsc{fft} to implement fast convolution, requiring only $\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a time domain search for compact binary coalescence.  We
will give a truly zero latency algorithm that competes in terms of floating point
operations per second with the conventional frequency domain method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of orthogonal
filters with far fewer coefficients than the original templates.  The reduction
in the number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples on
%
\editorial{Where do these numbers come from?}
%
average and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate on average.  This amounts to more than a factor of 100 in the
floating point operations per second required.  However, the resulting filters
are still not orthogonal across the mass parameter space, and are in fact
highly redundant.  We use the singular value decomposition to produce an
orthogonal filter set from the time sliced templates~\cite{Cannon:2010p10398}.
We find that this reduces the number of sample points that need to be filtered
by another factor of $\sim 100$.  The combined methods reduce the number of
floating point operations to the level where they are competitive with the
conventional matched filter approach.  In the remainder of this section we
describe the \lloid\ algorithm in detail and provide some basic computational
cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\textrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the convolution at a lower sample
rate.  Specifically we consider here time slice boundaries with the smallest
power-of-two sample rate that subcritically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}

@@

Definition of decimated detector data:
\begin{equation}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}

\editorial{Chad: This whole subsection needs a lot of work.
I just dropped our new notation in here.}%
In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decompostion in \eqref{eq:decomp} in the remainder
of this section.

In the next section we compute the expected computational cost scaling of this
decomposition and compare it with the brute force time domain implementation of
\eqref{eq:SNRTD} and higher latency frequency domain methods.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time domain or frequency domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in time slice $s$ \\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

\Table{\label{table:flops}Minimum computational cost of the \textsc{fir} filter bank, \textsc{fft} convolution, and \lloid\ methods for a bank of 657 templates sampled at 4096 Hz.}
\br
Method & \flops \\
\mr
\textsc{fir} & $2.4\times10^{13}$ \\
\textsc{fft} convolution & $2.6\times10^8$ \\
\lloid & $4.7\times10^8$ \\
\br
\end{tabular}
\end{indented}
\end{table}

\subsubsection{Conventional time domain method}

The conventional time domain method consists of a bank of \textsc{fir} filters, or sliding window dot products.  If there are $\numtmps$ templates, each $\tmpsamps$ samples in length, then each filter requires $M N$ multiplications and additions per sample, or $2 \numtmps \tmpsamps f^0$ floating point operations per second (\flops) at a sample rate $f^0$.

\subsubsection{Conventional frequency domain method}

The most common frequency domain method is known as the \emph{overlap-save} algorithm, described in \cite{numerical-recipes-chapter-13}.  It entails splitting the input into blocks of $D$ samples, $D > \tmpsamps$, each block overlapping the previous one by $D - \tmpsamps$ samples.  For each block, the algorithm computes the forward \fft\ of the data and the templates, multiplies them, and then computes the reverse \fft.

@@

The story is similar for the interpolation filters used to change the sample rates of the partial \textsc{snr} streams.  Interpolation of a data stream from a sample rate $f^s$ to $f^{s-1}$ consists of inserting zeros between the samples of the original stream, and then applying a lowpass filter with $N^\shortdownarrow$ coefficients.  The lowpass filter requires $2 M N^\shortdownarrow f^{s-1}$ \flops.  However, by taking advantage of the fact that by construction a fraction $f^{s-1}/f$ of the samples are zero, it is possible to build an efficient interpolator that requires only $M N^\shortuparrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 M N^\shortuparrow f^s$ \flops.

>>>>

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in a
search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 science runs.
We combine this with the orthogonal decomposition described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain as an \fir{} filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point operations per sample.  However, it is typically much more computationally efficient to use the convolution theorem and the \fft{} to implement fast convolution, requiring only $\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a time-domain search for compact binary coalescence.  We
will give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional frequency-domain method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of orthogonal
filters with far fewer coefficients than the original templates.  The reduction
in the number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time-domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples on
%
\editorial{Where do these numbers come from?}
%
average and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate on average.  This amounts to more than a factor of 100 in the
floating point operations per second required.  However, the resulting filters
are still not orthogonal across the mass parameter space, and are in fact
highly redundant.  We use the singular value decomposition to produce an
orthogonal filter set from the time sliced templates~\cite{Cannon:2010p10398}.
We find that this reduces the number of sample points that need to be filtered
by another factor of $\sim 100$.  The combined methods reduce the number of
floating point operations to the level where they are competitive with the
conventional matched filter approach.  In the remainder of this section we
describe the \lloid\ algorithm in detail and provide some basic computational
cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\mathrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the convolution at a lower sample
rate.  Specifically we consider here time slice boundaries with the smallest
power-of-two sample rate that sub-critically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time-sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}

@@

Definition of decimated detector data:
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k]
\end{equation}

\editorial{Chad: This whole subsection needs a lot of work.
I just dropped our new notation in here.}%
In order to minimize latency we propose using the time-domain convolution
presented in \eqref{eq:SNRTD}.  However, because the brute-force evaluation of
\eqref{eq:SNRTD} is far too costly to be useful, we will consider an
approximation to \eqref{eq:SNRTD} that can reduce substantially the cost of
\realtime\ filtering. This approximation has the form
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
where $u_{jk}(\tau)$ is an orthogonal basis set of filters spanning the space
of $\{x_i(\tau)\}$ and $\sigma_{jk} v_{ijk}$ is a tensor relating the filters
$u_{jk}(\tau)$ to the original filter set $\{x_i(\tau)\}$.  We claim that with
a suitable choice of filters $u_{jk}(\tau)$ one can reduce the computational
cost of \eqref{eq:SNRTD} sufficiently to feasibly search for
gravitational waves from compact binary coalescence in \realtime.  This
requires 1) exploiting
the time-frequency characteristics of the binary waveforms and 2) exploiting the redundancy of the template bank. We describe our
procedure for producing the decomposition in \eqref{eq:decomp} in the remainder
of this section.

In the next section we compute the expected computational cost scaling of this
decomposition and compare it with the brute-force time-domain implementation of
\eqref{eq:SNRTD} and higher latency frequency-domain methods.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time-domain or frequency-domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in time slice $s$ \\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

\Table{\label{table:flops}Minimum computational cost of the \fir{} filter bank, \fft{} convolution, and \lloid\ methods for a bank of 657 templates sampled at 4096 Hz.}
\br
Method & \flops \\
\mr
\fir{} & $2.4\times10^{13}$ \\
\fft{} convolution & $2.6\times10^8$ \\
\lloid & $4.7\times10^8$ \\
\br
\end{tabular}
\end{indented}
\end{table}

\subsubsection{Conventional time-domain method}

The conventional time-domain method consists of a bank of \fir{} filters, or sliding-window dot products.  If there are $\numtmps$ templates, each $\tmpsamps$ samples in length, then each filter requires $M N$ multiplications and additions per sample, or $2 \numtmps \tmpsamps f^0$ floating point operations per second (\flops) at a sample rate $f^0$.

\subsubsection{Conventional frequency-domain method}

The most common frequency-domain method is known as the \emph{overlap-save} algorithm, described in \cite{numerical-recipes-chapter-13}.  It entails splitting the input into blocks of $D$ samples, $D > \tmpsamps$, each block overlapping the previous one by $D - \tmpsamps$ samples.  For each block, the algorithm computes the forward \fft\ of the data and the templates, multiplies them, and then computes the reverse \fft.

@@

The story is similar for the interpolation filters used to change the sample rates of the partial \SNR{} streams.  Interpolation of a data stream from a sample rate $f^s$ to $f^{s-1}$ consists of inserting zeros between the samples of the original stream, and then applying a low-pass filter with $N^\shortdownarrow$ coefficients.  The low-pass filter requires $2 M N^\shortdownarrow f^{s-1}$ \flops.  However, by taking advantage of the fact that by construction a fraction $f^{s-1}/f$ of the samples are zero, it is possible to build an efficient interpolator that requires only $M N^\shortuparrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 M N^\shortuparrow f^s$ \flops.
